Draft Report

### 2020-11-18 DraftReport.Rmd
Updated draft report with regression table analysis and updated conclusion

---
title: "Draft Report"
author: "203 Section 6, Team 1"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

@@

For this report, the investigation will be centered around the effect of population metrics such as density and demographics and policy on the COVID case rates across the United States (US), which is grouped by the 50 states plus the District of Columbia (D.C.). This research question aims to measure the case rate per 100,000 residents within each state based on the make-up of the population, as well as the policy decisions that were or were not put into place in order to combat the rise of COVID cases. In this sense, the aim is to measure how dependent the COVID case rate is on that of which cannot be controlled (i.e., population and demographics) vs that of which can be controlled to a certain degree (i.e., implementation of policies to attempt to combat cases). With this information, one could suggest whether or not the proliferation of COVID can be controlled, and to the level at which these policies help with this control.

@@

Finally, with regard to the policy variables, this report focuses solely on mask mandates. Given this, it does not capture the effect of other policies that may or may not have a greater effect on the case rate (e.g., implementation of stay-at-home orders). This follows for the population-related variables, as they capture race/ethnicity and poverty, but not other characteristics such as proportions of residents with pre-existing conditions. In justification, these choices were largely made due to there either being an appropriate amount of samples in each category (e.g., most states have implemented basic policies such as the closure of non-essential businesses and the implementation of stay-at-home orders, and thus these were not included given the lack of samples for the states that have not implemented them), or that data on other policies or demographic measures were simply not readily available in the dataset.

@@

#### 2-b) Percentage living under the federal poverty line(2018)

@@

```{r}
ggplot(data = df,
  mapping = aes(x= poverty_pct))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(6,21,1)) +
  labs(title =  "Histogram of Poverty Percentage", x = "Percentage living under poverty line(2018)", y = 'Count')+
  scale_x_continuous(breaks=seq(6, 21, 1))
```
#### 2-c) Unemployed Percentage (2018)
Although, the distribution is of high peak in the middle of the range. Overall, it is not highly skewed or heavily tailed. In addition, the unemployed rate is correlated to the variable to the poverty rate, which is in an agreement with intuitive expectation.

@@

#### 2-d) Seniors Percentage(Age 65+)
The seniors percentage (age 65+) is of a nearly normal distribution.

```{r}
summary(df$senior_pct)
```
```{r}
ggplot(data = df,
  mapping = aes(x= senior_pct))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(0.10,0.22,0.01)) +
  labs(title =  "Histogram of Seniors Percentage (65+)", x = "Seniors percentage", y = 'Count')+
  scale_x_continuous(breaks=seq(0.1, 0.25, 0.05))
```

@@

```{r}
df %>% 
  ggplot(aes(white_population_pct, case_rate_100k, color = log10(population_density))) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate Per 100000 to White Population Percentage', 
    x = 'White percentage of total population', 
    y = 'Case rate per 100000',
    color = 'Log 10 of population density'
  )
```
```{r}
df %>% 
  ggplot(aes(x = white_population_pct, y = case_rate_100k, color = poverty_pct)) + 
  geom_point() +
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate per 100000 to White Population Percentage', 
    x = 'White percentage of total population', 
    y = 'Case rate per 100000',
    color = 'Poverty percentage'
  )
```
```{r}
df %>% 
  ggplot(aes(x = poverty_pct , y = case_rate_100k, color = log10(population_density))) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate per 100000 to Poverty Percentage', 
    x = 'Poverty Percentage', 
    y = 'Case rate per 100000',
    color = 'Log 10 of population density'
  )
```

@@

#### Model 2-a
$$
\begin{aligned}
 Case.Rate.Per.100000 =\beta_0 &+\beta_1 log(Population.Density.Per.Square.Miles)\\&+\beta_2(White.Percentage.of.Total.Population)
\end{aligned}
$$

@@

Firstly, "mask_public_bool" column is created from "mask_public", with 1 represents that the state has a pulbic mask policy, and 0 means there's no explicit public mask policy from the state.
```{r}

@@

Histograms below that shows the distribution of states that has public mask policy or not. Both distributions exhibits certain degree of normality. 
```{r}
hist(df$case_rate_100k[df$mask_public_bool == 1], xlab='Case Rate', ylab='Count', main='Case Rate - Mask Mandate in Public')
hist(df$case_rate_100k[df$mask_public_bool == 0], xlab='Case Rate', ylab='Count', main='Case Rate  - No Mask Mandate in Public')
```

A plot that shows the relationship between different policies vs case rate.
```{r}
df %>% 
  ggplot(aes(x = mask_public_bool, y = case_rate_100k, color = case_rate_100k)) + 
  geom_point(aes(size=case_rate_100k)) + 
  labs(
    title = 'Case Rate vs Mask Public Mandate', 
    x = 'Mask Mandate in Public', 
    y = 'Case Rate', 
    color = 'Case Rate'
  )
```

#### Model 3-a
Next, the linear model is created from model 2 with the mask policy variables added for the regression. The regression coefficient shows that Mask_public_bool variable is highly significant. The result suggests that public mask mandate policy has postive effect in bringing the case number down.

@@

#### Model 3-b&c
Considering that there could be multiple policy variables that are relevant to predicting the case number. We have also examined the "Closed.other.non.essential.businesses" variable (renamed as "business_closed") which represents the date when the state issued closure of non-essential businesses. As "business_closed" is a date variable, a transformation was performed to convert the date into a variable "business_close_diff" which represents the days difference between the business close date of the state compared to the average date of all the states.
```{r}

df$business_close_diff =as.numeric(df$business_closed-mean(df$business_closed, na.rm=TRUE))
head(df)
```

From the plot of business close date difference against case rate, the graph shows that there is some degree of linearity between the two variables.
```{r warning=FALSE}
df %>% 
  ggplot(aes(x = business_close_diff, y = case_rate_100k, color = business_close_diff)) + 
  geom_point() + 
  geom_smooth(se = FALSE) +
  labs(
    title = 'Case Rate vs Non-Essential Business Close Date difference compared to Average', 
    x = 'Non-Essential Business Close Date difference compared to Average', 
    y = 'Case Rate', 
    color = 'Case Rate'
  )
```

@@

However, the regression with both policy variables "mask_pubic_bool" and "business_close_diff" shows that neither the the variable is significant. 
```{r}
model3c <- lm(case_rate_100k ~ business_close_diff + mask_public_bool, data = df)
coeftest(model3c, vcov = vcovHC)
```

@@

Secondly, the process being modeled must be described by a linear conditional expectation function of the variables which are included, in order for the linear regression to be valid. This can be assessed after the fact by visualizing the model over the range of each variable alongside the data, or also by visualizing the residuals over the range of each variable. For the purpose of this assumption and other model or data related assumptions, only the Model 3A will be used as it is the most inclusive. 
```{r}
avg_pop <- log10(mean(df$population_density))
avg_pov <- mean(df$poverty_pct)
avg_mask <- mean(df$mask_public_bool)
y_pop <- model3a$coefficients[1] + model3a$coefficients[2]*log10(df$population_density) + model3a$coefficients[3]*avg_pov + model3a$coefficients[4]*avg_mask
y_pov <- model3a$coefficients[1] + model3a$coefficients[2]*avg_pop + model3a$coefficients[3]*df$poverty_pct + model3a$coefficients[4]*avg_mask
y_pop_nomask <- model3a$coefficients[1] + model3a$coefficients[2]*avg_pop + model3a$coefficients[3]*avg_pov
y_pov_nomask <- model3a$coefficients[1] + model3a$coefficients[2]*avg_pop + model3a$coefficients[3]*df$poverty_pct

df %>% 
  ggplot() + 
  geom_point(aes(x = log10(population_density), y = case_rate_100k, group = factor(mask_public_bool), color = factor(mask_public_bool))) + 
  geom_line(aes(x = log10(population_density), y = y_pop, color = "red")) +
  geom_line(aes(x = log10(population_density), y = y_pop_nomask, color = "blue")) +
  scale_color_manual(labels = c("No", "Yes", "No", "Yes"), values = c("blue", "red", "blue", "red")) +
  labs(
    title = 'Linearity of Variables, Population Density', 
    x = 'Base 10 Log of Population Density', 
    y = 'Case Rate', 
    color = 'Mask Mandate'
  )
df %>% 
  ggplot() + 
  geom_point(aes(x = poverty_pct, y = case_rate_100k, group = factor(mask_public_bool), color = factor(mask_public_bool))) + 
  geom_line(aes(x = poverty_pct, y = y_pov, color = "red")) +
  geom_line(aes(x = poverty_pct, y = y_pov_nomask, color = "blue")) +
  scale_color_manual(labels = c("No", "Yes", "No", "Yes"), values = c("blue", "red", "blue", "red")) +
  labs(
    title = 'Linearity of Variables, Poverty Percentage', 
    x = 'Poverty Percentage', 
    y = 'Case Rate', 
    color = 'Mask Mandate'
  )
```
Linearity for the three regressed variables is demonstrated above. As can be seen when mask policy and population density are varied, the data generally follows a linear trend. This is similar when the poverty percentage is varied, although at higher levels of poverty the presence of a mask mandate made the case rate unexpectedly low. It appears that possible a negative quadratice function may fit the data better, although it appears close enough to linear that this assumption may hold.

@@

@@

- Discuss variables included and not included in the model
- Talk about Adjusted R2
- Talk about significant variables in the table
- Provide a statement that the table may change in the final report with more variables and maybe different key variables (case rate and population density)

@@

Conversely, the hypotheses that population and demographics would have an effect on the COVID case rate was unable to be accepted. Although upon investigating the poverty line percentage variable in model 2 there was a detection of notable adjacency to a significant result with a coefficient value of 0.06286, it was not enough to be truly significant and reject the null hypothesis. Similarly, the relationship of race/ethnicity as well as age on case rate was also insignificant with no apparent correlation.

Given these results in conjunction with the original aim of the research question, it can be suggested that variables that can be controlled (in this case, mask policy) has a greater effect on stymieing the spread of COVID-19 than the variables that may be inherent to a state, such as population density and age/ethnicity demographics. This is a promising insight, as it provides opportunity for states to implement and apply policies to intervene and control the case rate without being worried about its varying effects depending on population characteristics.

>>>>

---
title: "Draft Report"
author: "203 Section 6, Team 1"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

@@

For this report, the investigation will be centered around the effect of population metrics such as density and demographics and policy on the COVID case rates across the United States (US), which is grouped by the 50 states plus the District of Columbia (D.C.). This research question aims to measure the case rate per 100,000 residents within each state based on the make-up of the population, as well as the policy decisions that were or were not put into place in order to combat the rise of COVID cases. In this sense, the aim is to measure how dependent the COVID case rate is on that of which cannot be controlled (i.e., population and demographics) vs that of which can be controlled to a certain degree (i.e., implementation of policies to attempt to combat case rate). With this information, one could suggest whether or not the proliferation of COVID can be controlled, and to the level at which these policies help with this control.

@@

Finally, with regard to the policy variables, this report focuses mainly on mask mandates. Given this, it does not capture the effect of other policies that may or may not have a greater effect on the case rate (e.g., implementation of stay-at-home orders). This follows for the population-related variables, as they capture race/ethnicity and poverty, but not other characteristics such as proportions of residents with pre-existing conditions. In justification, these choices were largely made due to there either being an appropriate amount of samples in each category (e.g., most states have implemented basic policies such as the closure of non-essential businesses and the implementation of stay-at-home orders, and thus these were not included given the lack of samples for the states that have not implemented them), or that data on other policies or demographic measures were simply not readily available in the dataset.

@@

#### 2-b) Percentage living under the federal poverty line (2018)

@@

```{r}
ggplot(data = df,
  mapping = aes(x= poverty_pct))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(6,21,1)) +
  labs(title =  "Histogram of Poverty Percentage", x = "Percentage living under poverty line(2018)", y = 'Count')+
  scale_x_continuous(breaks=seq(6, 21, 1))
```

#### 2-c) Unemployed Percentage (2018)

Although, the distribution is of high peak in the middle of the range. Overall, it is not highly skewed or heavily tailed. In addition, the unemployed rate is correlated to the variable to the poverty rate, which is in an agreement with intuitive expectation.

@@

#### 2-d) Seniors Percentage (Age 65+)

The seniors percentage (age 65+) is of a nearly normal distribution.

```{r}
summary(df$senior_pct)
```

```{r}
ggplot(data = df,
  mapping = aes(x= senior_pct))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(0.10,0.22,0.01)) +
  labs(title =  "Histogram of Seniors Percentage (65+)", x = "Seniors percentage", y = 'Count')+
  scale_x_continuous(breaks=seq(0.1, 0.25, 0.05))
```

@@

```{r}
df %>% 
  ggplot(aes(white_population_pct, case_rate_100k, color = log10(population_density))) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate Per 100000 to White Population Percentage', 
    x = 'White percentage of total population', 
    y = 'Case rate per 100000',
    color = 'Log 10 of population density'
  )
```

```{r}
df %>% 
  ggplot(aes(x = white_population_pct, y = case_rate_100k, color = poverty_pct)) + 
  geom_point() +
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate per 100000 to White Population Percentage', 
    x = 'White percentage of total population', 
    y = 'Case rate per 100000',
    color = 'Poverty percentage'
  )
```

```{r}
df %>% 
  ggplot(aes(x = poverty_pct , y = case_rate_100k, color = log10(population_density))) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'Relation of Case Rate per 100000 to Poverty Percentage', 
    x = 'Poverty Percentage', 
    y = 'Case rate per 100000',
    color = 'Log 10 of population density'
  )
```

@@

#### Model 2-a

$$
\begin{aligned}
 Case.Rate.Per.100000 =\beta_0 &+\beta_1 log(Population.Density.Per.Square.Miles)\\&+\beta_2(White.Percentage.of.Total.Population)
\end{aligned}
$$

@@

Firstly, "mask_public_bool" column is created from "mask_public", with 1 represents that the state has a pulbic mask policy, and 0 means there's no explicit public mask policy from the state.

```{r}

@@

Histograms below that shows the distribution of states that has public mask policy or not. Both distributions exhibits certain degree of normality.

```{r}
hist(df$case_rate_100k[df$mask_public_bool == 1], xlab='Case Rate', ylab='Count', main='Case Rate - Mask Mandate in Public')
hist(df$case_rate_100k[df$mask_public_bool == 0], xlab='Case Rate', ylab='Count', main='Case Rate  - No Mask Mandate in Public')
```

A plot that shows the relationship between different policies vs case rate.

```{r}
df %>% 
  ggplot(aes(x = mask_public_bool, y = case_rate_100k, color = case_rate_100k)) + 
  geom_point(aes(size=case_rate_100k)) + 
  labs(
    title = 'Case Rate vs Mask Public Mandate', 
    x = 'Mask Mandate in Public', 
    y = 'Case Rate', 
    color = 'Case Rate'
  )
```

#### Model 3-a

Next, the linear model is created from model 2 with the mask policy variables added for the regression. The regression coefficient shows that Mask_public_bool variable is highly significant. The result suggests that public mask mandate policy has postive effect in bringing the case number down.

@@

#### Model 3-b&c

Considering that there could be multiple policy variables that are relevant to predicting the case number. We have also examined the "Closed.other.non.essential.businesses" variable (renamed as "business_closed") which represents the date when the state issued closure of non-essential businesses. As "business_closed" is a date variable, a transformation was performed to convert the date into a variable "business_close_diff" which represents the days difference between the business close date of the state compared to the average date of all the states.

```{r}
df$business_close_diff =as.numeric(df$business_closed-mean(df$business_closed, na.rm=TRUE))
head(df)
```

From the plot of business close date difference against case rate, the graph shows that there is some degree of linearity between the two variables.

```{r warning=FALSE}
df %>% 
  ggplot(aes(x = business_close_diff, y = case_rate_100k, color = business_close_diff)) + 
  geom_point() + 
  geom_smooth(se = FALSE) +
  labs(
    title = 'Case Rate vs Non-Essential Business Close Date difference compared to Average', 
    x = 'Non-Essential Business Close Date difference compared to Average', 
    y = 'Case Rate', 
    color = 'Case Rate'
  )
```

@@

However, the regression with both policy variables "mask_pubic_bool" and "business_close_diff" shows that neither the the variable is significant.

```{r}
model3c <- lm(case_rate_100k ~ business_close_diff + mask_public_bool, data = df)
coeftest(model3c, vcov = vcovHC)
```

@@

Secondly, the process being modeled must be described by a linear conditional expectation function of the variables which are included, in order for the linear regression to be valid. This can be assessed after the fact by visualizing the model over the range of each variable alongside the data, or also by visualizing the residuals over the range of each variable. For the purpose of this assumption and other model or data related assumptions, only the Model 3A will be used as it is the most inclusive.

```{r}
avg_pop <- log10(mean(df$population_density))
avg_pov <- mean(df$poverty_pct)
avg_mask <- mean(df$mask_public_bool)
y_pop <- model3a$coefficients[1] + model3a$coefficients[2]*log10(df$population_density) + model3a$coefficients[3]*avg_pov + model3a$coefficients[4]*avg_mask
y_pov <- model3a$coefficients[1] + model3a$coefficients[2]*avg_pop + model3a$coefficients[3]*df$poverty_pct + model3a$coefficients[4]*avg_mask
y_pop_nomask <- model3a$coefficients[1] + model3a$coefficients[2]*avg_pop + model3a$coefficients[3]*avg_pov
y_pov_nomask <- model3a$coefficients[1] + model3a$coefficients[2]*avg_pop + model3a$coefficients[3]*df$poverty_pct

df %>% 
  ggplot() + 
  geom_point(aes(x = log10(population_density), y = case_rate_100k, group = factor(mask_public_bool), color = factor(mask_public_bool))) + 
  geom_line(aes(x = log10(population_density), y = y_pop, color = "red")) +
  geom_line(aes(x = log10(population_density), y = y_pop_nomask, color = "blue")) +
  scale_color_manual(labels = c("No", "Yes", "No", "Yes"), values = c("blue", "red", "blue", "red")) +
  labs(
    title = 'Linearity of Variables, Population Density', 
    x = 'Base 10 Log of Population Density', 
    y = 'Case Rate', 
    color = 'Mask Mandate'
  )
df %>% 
  ggplot() + 
  geom_point(aes(x = poverty_pct, y = case_rate_100k, group = factor(mask_public_bool), color = factor(mask_public_bool))) + 
  geom_line(aes(x = poverty_pct, y = y_pov, color = "red")) +
  geom_line(aes(x = poverty_pct, y = y_pov_nomask, color = "blue")) +
  scale_color_manual(labels = c("No", "Yes", "No", "Yes"), values = c("blue", "red", "blue", "red")) +
  labs(
    title = 'Linearity of Variables, Poverty Percentage', 
    x = 'Poverty Percentage', 
    y = 'Case Rate', 
    color = 'Mask Mandate'
  )
```

Linearity for the three regressed variables is demonstrated above. As can be seen when mask policy and population density are varied, the data generally follows a linear trend. This is similar when the poverty percentage is varied, although at higher levels of poverty the presence of a mask mandate made the case rate unexpectedly low. It appears that possible a negative quadratice function may fit the data better, although it appears close enough to linear that this assumption may hold.

@@

Although there were several versions of models created for model 2 and 3, through the analysis of the appropriate variables respective to each model a single model version was selected to be included in the regression table. This was determined based on factors of significance when investigating the effect relationship of these variables on the outcome variable, which is detailed in section 2.

For the regression table below, model 1, model 2b, and model 3a were selected as the most significant models to include.

@@

When observing the table above, three elements stand out.

Firstly, there is a difference in the calculated standard error between the coefficient test run on the linear models and the regression table. Taking a look at the standard error for the poverty percentage coefficient, the value in the table returns a significant result with a standard error of 55.125, as opposed to the insignificant result of a standard error of 57.684. This follows for model 3, with the table showing a significant result with standard error of 50.872, as opposed to an insignificant result of a standard error of 59.583.

Secondly, following the first observation, when designating the poverty percentage variable as significant, the regression table suggests that both covariate variables (including the mask mandate variable) are indeed significant with the COVID case rate. For the poverty percentage effect, the coefficient highlights that there is a positive correlation between the poverty percentage and case rate (as poverty percentage increases, so does case rate). The practical significance of this shows that for every percentage increase of population below the federal poverty line, there is an increase of roughly 112 cases per 100k for model 2, and roughly 104 cases per 100k for model 3. As for the mask mandate variable, the coeeficient highlights a negative correlation between it and the case rate (if the mask mandate is implemented, the case rate decreases). The practical significance is such that if the mask mandate has been implemented, the case rate per 100k will decrease by roughly 1000 cases.

Thirdly, we notice that the R2 value increases by a large value for every additional variable that is included within the model. Initially, we start with a very low R2 value, suggesting that almost no variance is explained in model 1. However, although still low, there is close to a 9 times increase in the amount of variance that is explained in model 2. Finally, in model 3, the R2 value increases by roughly 2.75 times the previous value, and a decent amount of the variance can be seen to be explained by the model. The adjusted R2 that accounts for the increase in number of variable terms in the model also correspondingly increases as the covariates are added.

As a note, it should be stressed that the models and variables included in this regression table may change when the final report is created to provide a more descriptive analysis of the original research question.

@@

Conversely, the hypotheses that population and demographics would have an effect on the COVID case rate was unable to be accepted. Although upon investigating the poverty line percentage variable in model 2 there was a detection of notable adjacency to a significant result (and an actual significant result when using the `r stargazer` to compute the regression table) with a coefficient value of 0.06286, it was not enough to be truly significant and reject the null hypothesis. Similarly, the relationship of race/ethnicity as well as age on case rate was also insignificant with no apparent correlation.

Given these results in conjunction with the original aim of the research question, it can be suggested that variables that can be controlled (in this case, mask policy) has a greater effect on stymieing the spread of COVID-19 than the variables that may be inherent to a state, such as population density and age/ethnicity demographics. This is a promising insight, as it provides opportunity for states to implement and apply policies to intervene and control the case rate without being worried about its varying effects depending on population characteristics. However, it should once again be stressed that the final report may include and analyze alternative variables and apply them to different models based on the findings from this draft report, which as a result may also lead to more concrete or slightly varied insights.
